---
title: "Atomic Data Types"
---

{{< include _static/_shared.qmd >}}

## Exposition

### Introduction

In this exercise we will explore three important atomic data types called
numeric, logical, and character. We call them *atomic* because they can
not be broken down into simpler data structures.[^notatomic]

[^notatomic]: Much like we now know that atoms can, in fact, be split into smaller
subatomic particles, these atomic data types can also be split further, but that
is far beyond the scope of this course.

Contrast this with *complex* data types that we will learn about in
later lessons and which are composed by these atomic data types.

### Numeric

You have already met the numeric data type in [Rithmatic](01-Rithmatic.html).
How do you know? We can just ask R if `1`{.r} is numeric by
typing and running `is.numeric(1)`{.r}. Try it yourself:

```{webr}
#| exercise: e1
```

::: { .solution exercise="e1" }
```{webr}
#| exercise: e1
#| solution: true
is.numeric(1)
```
:::

```{webr}
#| exercise: e1
#| check: true
gradethis::grade_this_code()
```

`is.numeric`{.r} is a name that refers to a method or recipe in R, called a *function*, that determines if the value it recieves is a number or not, i.e., is it numeric or not.

Now, try running `is.numeric(2 + 2)`{.r} to see the result.

```{webr}
#| exercise: e2
```

```{webr}
#| exercise: e2
#| solution: true
is.numeric(2 + 2)
```

```{webr}
#| exercise: e2
#| check: true
gradethis::grade_this_code()
```

### Logical

I hope you've noticed that R responds `TRUE`{.r} when the value the `is.numeric`{.r} function recieves is indeed numeric.

The value `TRUE`{.r} is a kind of the next atomic data type we will explore, i.e., the logical data type.

`TRUE`{.r} is R's word for "yes." R is case-sensitive, so `TRUE`{.r} is not the same as `true`{.r}, `True`{.r},
or even `TrUe`{.r}, so be sure to type it carefully.
R can be very particular about such things.

`TRUE`{.r} is not a numeric value.  How can you find out?

```{webr}
#| exercise: e3
```

```{webr}
#| exercise: e3
#| solution: true
is.numeric(TRUE)
```

```{webr}
#| exercise: e3
#| check: true
gradethis::grade_this_code()
```

Instead of `TRUE`{.r}, R responds with `FALSE`{.r}, which is R's word for "no."  `TRUE`{.r} and `FALSE`{.r} are the only two possible values for the logical data type in R.

Can you guess how we test for logical values? Try running `is.logical(FALSE)`{.r}.

```{webr}
#| exercise: e4
```

```{webr}
#| exercise: e4
#| solution: true
is.logical(FALSE)
```

```{webr}
#| exercise: e4
#| check: true
gradethis::grade_this_code()
```

If you ran `is.logical(TRUE)`{.r}, what should the answer be?
Type and check it below.

```{webr}
#| exercise: e5
```

```{webr}
#| exercise: e5
#| solution: true
TRUE
```

```{webr}
#| exercise: e5
#| check: true
gradethis::grade_this_code()
```

R will let you use `T`{.r} for `TRUE`{.r} and `F`{.r} for `FALSE`{.r}, but please don't do that in this class or
in your work.  It makes your code harder to read for others, just as it made it harder for your 5th grade teacher to
grade your true-false test. However, I want you to know that in case you see someone use it in their code or if your instructor is being lazy as he types in class. After all,
that's only what it should be---a shortcut. Note, though your instructor always types `TRUE`{.r} and `FALSE`{.r} out completely in his production code as 4--5 extra characters are worth the clarity.

OK, I'm pretty confident you have this now. What would `is.logical(1)`{.r} return?

```{webr}
#| exercise: e6
```

```{webr}
#| exercise: e6
#| solution: true
FALSE
```

```{webr}
#| exercise: e6
#| check: true
gradethis::grade_this_code()
```

### Character

The last important atomic data type you should know about
for now is the character data type. The character type is used to represent text in R. So you know some of the lingo text for computers is often called *character strings* or simply *strings*.

R knows you intend something to be a string if you enclose it in quotes. The preferred style in R code is to use double quotes,

```{r}
#| eval: false
"Hello!"
```

but single quotes

```{r}
#| eval: false
'Hello!'
```

also work just fine and are sometimes important for getting
your code to work right as we will see.

See what R does when you type and run `"Hello, my name is R"`{.r}.

```{webr}
#| exercise: e7
```

```{webr}
#| exercise: e7
#| solution: true
"Hello, my name is R"
```

```{webr}
#| exercise: e7
#| check: true
gradethis::grade_this({
  if (.user_code == '"Hello, my name is R"') {
    pass()
  } else if (.user_code == "'Hello, my name is R'") {
    pass(glue::glue("It works, but remember double quotes are preferred",
     " when writing R code. Note that even R's response uses them!"
    ))
  } else {
    fail("Type it just like I wrote it.", encourage = TRUE)
  }
})
```

Even though double quotes are preferred, single quotes can be used. Let's
test that by typing and running `'What is your name?'`{.r}.

```{webr}
#| exercise: e8
```

```{webr}
#| exercise: e8
#| solution: true
'What is your name?'
```

```{webr}
#| exercise: e8
#| check: true
gradethis::grade_this({
  if (.user_code == "'What is your name?'") {
    pass()
  } else if (.user_code == '"What is your name?"') {
    pass(glue::glue("You are right that double quotes are preferred,",
     " but this time I wanted you to use single quotes."
    ))
  } else {
    fail("Type it just like I wrote it.", encourage = TRUE)
  }
})
```

```{adm}
#| exercise: e8
#| type: open-on-pass
Look at R's response above. R said back `"What is your name?"` with double
quotes even when single quotes were used, which emphasizes, in part, why double
quotes are preferred when writing R code.
```

Since the `is.logical`{.r} function tests whether something is of the logical
datatype, and `is.numeric`{.r} tests whether something is of the numeric data
type, what function do you think tests whether something is of the character
type?

Go for it!  Test whether

```{r}
#| eval: false
"Hi!"
```

is of type character.

```{webr}
#| exercise: e9
```

```{webr}
#| exercise: e9
#| solution: true
is.character("Hi!")
```

```{webr}
#| exercise: e9
#| check: true
gradethis::grade_this_code()
```

Now it is time to try out your skills. Experiment below and figure out
which atomic data type (numeric, logical, or character) is the result of
dividing another number by zero.

```{webr}
#| caption: Experiment
```
